---
title: "Data Classes"
author: "APH101 Daiyun Huang"
output:
  ioslides_presentation:
    css: ./styles.css
    widescreen: yes
---

```{r, echo = FALSE, message=FALSE}
# library(dplyr)
suppressPackageStartupMessages(library(dplyr))
library(readr)
library(forcats)
```

## Acknowledgement

Thanks to Dr. John Muschelli and Dr. Andrew Jaffe for sharing their course materials online. This material is prepared based on their work.

## Contents

In this section, we will discuss the different (one dimensional/vector) data types/classes in R

* numeric
* character
* integer
* factor
* logical
* Date/POSIXct

as well as the other more complex R classes

* lists
* data.frame
* matrix

## Data Types

* One dimensional types ('vectors'):
    * Character: strings or individual characters, quoted
    * Numeric: any real number(s)
    * Integer: any integer(s)/whole numbers
    * Factor: categorical/qualitative variables
    * Logical: variables composed of TRUE or FALSE
    * Date/POSIXct: represents calendar dates and times

## Character and numeric

`character` and `numeric` types.

```{r numChar}
class(c("XJTLU", "AoP"))
class(c(1, 3, 6))
```

## Integer

`Integer` is a special subset of `numeric` that contains only whole numbers

A sequence of numbers is an example of the integer type

```{r seq}
x = seq(from = 1, to = 5) # seq() is a function
x
class(x)
```

## Integer

The colon `:` is a shortcut for making sequences of numbers

`[num1]:[num2]` makes a consecutive integer sequence from `[num1]` to `[num2]` by 1

```{r seqShort}
1:10
```

If you want to make a sequence with interval other 1, try

```{r seqBy}
x = seq(from = 1, to = 10, by = 2) # seq() is a function
x
```

## Logical

`logical` is a type that only has two possible elements: `TRUE` and `FALSE`

```{r logical1}
x = c(TRUE, FALSE, TRUE, TRUE, FALSE)
class(x)
is.numeric(c("XJTLU", "AoP"))
is.character(c("XJTLU", "AoP"))
```

## Logical

Note that `logical` elements are NOT in quotes. 

```{r logical2}
z = c("TRUE", "FALSE", "TRUE", "FALSE")
class(z)
as.logical(z)
```

## Logical

In certain cases, it can be useful to treat `logical` values as `numeric`. In this case, `TRUE = 1` and `FALSE = 0`. `sum()` and `mean()` work on `logical` vectors - they return the total and proportion of `TRUE` elements, respectively.

```{r logical_z}
sum(as.logical(z))
mean(as.logical(z))
```

## General Class Information

There are two useful functions associated with practically all R classes, which relate to logically checking the underlying class (`is.CLASS_()`) and coercing between classes (`as.CLASS_()`).

```{r logical_coercion}
is.numeric(c("XJTLU", "AoP"))
is.character(c("XJTLU", "AoP"))
```

## General Class Information

There are two useful functions associated with practically all R classes, which relate to logically checking the underlying class (`is.CLASS_()`) and coercing between classes (`as.CLASS_()`).

```{r logical_coercion2}
as.character(c(1, 3, 6))
as.numeric(c("XJTLU", "AoP"))
```

## Factors

A `factor` is a special `character` vector where the elements have pre-defined groups or 'levels'. You can think of these as qualitative or categorical variables. Consider the following categorical variable `x` which records the gender for five children as either boy or girl.

```{r factor1}
x = factor(c("boy", "girl", "girl", "boy", "girl"))
x 
class(x)
```

Note that levels are, by default, in alphanumerical order.

## Factors

Factors are used to represent categorical data, and can also be used for ordinal data (i.e., categories have an intrinsic ordering) by setting `ordered = TRUE`.

Note that R reads in character strings as factors by default in functions like `read.csv()` (but not `read_csv`)

'The function factor is used to encode a vector as a factor (the terms 'category' and 'enumerated type' are also used for factors). If argument ordered is TRUE, the factor levels are assumed to be ordered.'

```
factor(x = character(), levels, labels = levels,
       exclude = NA, ordered = is.ordered(x))
```

## Factors

Since the order of the levels can matter, how can we alter or set the order of these levels in a way other than alphanumeric order?

```{r factor2}
cc = factor(c("case", "case", "case", "control", "control", "control"))
cc
```

We can reset the levels using the `levels` function, but this is **bad** and can cause problems. You should do this using the `levels` argument in the `factor()`

```{r}
levels(cc) = c("control", "case")
cc
```

## Factors

Note that the levels are alphabetically ordered by default. We can also specify the levels within the factor call

```{r factor_cc_again}
casecontrol = c("case", "case", "case", "control", "control", "control")
factor(casecontrol, levels = c("control", "case") )
factor(casecontrol, levels = c("control", "case"), ordered=TRUE)
```

## Factors

Another way to do this once you already have the factor made is with the `relevel()` function. 

```{r factorCheck}
cc = factor(c("case", "case", "case", "control", "control", "control"))
relevel(cc, "control")
```

## Factors

One of the core "tidyverse" packages is `forcats` which offers useful functionality for interacting with factors. For example, there is a function for releveling factors here:

```{r}
fct_relevel(cc, "control")
```

## Factors 

There are other useful functions for dictating the levels of factors, like in the order they appears in the vector, by frequency, or into collapsed groups.

```{r}
levels(fct_inorder(chickwts$feed))
levels(fct_infreq(chickwts$feed))
levels(fct_lump(chickwts$feed, n=1))
```

## Factors

Factors can be converted to `numeric` or `character` very easily

```{r factor3}
x = factor(casecontrol, levels = c("control", "case"))
as.character(x)
as.numeric(x)
```

## Creating categorical variables

The `rep()` ["repeat"] function is useful for creating new variables 

```{r rep1}
bg = rep(c("boy", "girl"),each=50)
head(bg)
bg2 = rep(c("boy", "girl"),times=50)
head(bg2)
length(bg) == length(bg2)
```

## Dates

Dates are represented as the number of days since 1970-01-01, with negative values for earlier dates.

```{r}
# use as.Date( ) to convert strings to dates
mydates <- as.Date(c("2023-04-07", "2023-01-01"))
mydates
# number of days between 4/7/23 and 1/1/23
days <- mydates[1] - mydates[2]
days
```

`Sys.Date()` returns today's date.
`date()` returns the current date and time.

## Dates

The following symbols can be used with the `format()` function to print dates.

* %d	day as a number (0-31)  01-31
* %a  abbreviated weekday  Mon
* %A	unabbreviated weekday  Monday
* %m	month (00-12)  00-12
* %b  abbreviated month  Jan
* %B	unabbreviated month  January
* %y  2-digit year  07
* %Y  4-digit year  2007

## Here is an example.

```{r, message = FALSE}
# print today's date
today <- Sys.Date()
format(today, format="%B %d %Y")
```
## Character to Date

You can use the `as.Date()` function to convert character data to dates. The format is `as.Date(x, "format")`, where `x` is the character data and format gives the appropriate format.

```{r}
# convert date info in format 'mm/dd/yyyy'
strDates <- c("01/01/2023", "04/07/2023")
dates <- as.Date(strDates, "%m/%d/%Y")
```

The default format is yyyy-mm-dd

```{r}
mydates <- as.Date(c("2023-01-01", "2023-04-07"))
```

## Date to Character

You can convert dates to character data using the `as.Character()` function.

```{r}
# convert dates to character data
strDates <- as.character(dates)
```

## Lubridate

`lubriate` is an R package that makes it easier to work with dates and times.

```{r}
library(lubridate)
```

The warning simply means that after importing the new library, the basic functions (date, intersect, setdiff, union) are replaced by new functions with the same name in the imported package.

## Lubridate

We can use date transformation function from `lubriate` that matches the date pattern in the strings, for example

* mdy : month day year
* ymd : year month day
* ydm : year day month
* ymd_hms : year month day hours minutes seconds

and so forth. For example, the following date is in the format year month day hours minutes seconds.

```{r}
x = c("2014-02-4 05:02:00","2016/09/24 14:02:00")
ymd_hms(x)
```

## Lubridate

If we use the wrong format, then we will get an error.
```{r}
ymd_hm(x)
```

## POSIXct

The `POSIXct` class is like a more general date format (with hours, minutes, seconds).

```{r, message = FALSE}
x = c("2023-01-01 05:02:00", "2023/04/07 14:02:00")
dates = ymd_hms(x)
class(dates)
```
## Adding Periods of time

The `as.Period` command is helpful for adding time to a date:

```{r}
current_time = Sys.time()
current_time
class(current_time)
current_time + as.period(20, unit = "minutes") # the future
```

## Differences in Times 

You can subtract times as well, the `difftime` function is helpful as you can set the units (note it does `time1 - time2`):

```{r}
the_future = ymd_hms("2023-12-31 11:59:59")
the_future - current_time
difftime(the_future, current_time, units = "weeks")
```

## Data Classes:

* Two dimensional classes:
    * `data.frame`: traditional 'Excel' spreadsheets
        * Each column can have a different class, from above
    * Matrix: two-dimensional data, composed of rows and columns. Unlike data frames, the entire matrix is composed of one R class, e.g. all numeric or all characters.

## Matrices

```{r matrix}
n = 1:9 
n
mat = matrix(n, nrow = 3)
mat
```

## Matrices

```{r matrix2}
x <- 1:4
x %*% x    # scalar ("inner") product (1 x 1 matrix)
```

```{r matrix3}
y <- diag(x)
y
```

## Matrices
```{r matrix4}
z <- matrix(1:12, ncol = 3, nrow = 4)
z
```

## Matrices

```{r matrix5}
y %*% z
y %*% x
x %*% z
```

## Eigenvalue and Eigenvector
```{r}
A <- matrix(c(13, -4, 2, -4, 11, -2, 2, -2, 8), 3, 3, byrow=TRUE)
A
```

```{r}
ev <- eigen(A)
(values <- ev$values)
(vectors <- ev$vectors)
```

## Data Selection

Matrices have two "slots" you can use to select data, which represent rows and columns, that are separated by a comma, so the syntax is `matrix[row,column]`. Note you cannot use `dplyr` functions on matrices.

```{r subset3}
mat[1, 1] # individual entry: row 1, column 1
mat[1, ] # first row
mat[, 1] # first columns
```

## Data Selection

Note that the class of the returned object is no longer a matrix

```{r subset4}
class(mat[1, ])
class(mat[, 1])
```

## Data Frames

To review, the `data.frame`/`tbl_df` are the other two dimensional variable classes. 

Again, data frames are like matrices, but each column is a vector that can have its own class. So some columns might be `character` and others might be `numeric`, while others maybe a `factor`.

## Data Frames

```{r}
# Create a data frame
Data_Frame <- data.frame (Training = c("Strength", "Stamina", "Other"),
                          Pulse = c(100, 150, 120),
                          Duration = c(60, 30, 45))
Data_Frame
summary(Data_Frame)
```

## Lists

* One other data type that is the most generic are `lists`.
* Can be created using list()
* Can hold vectors, strings, matrices, models, list of other list, lists upon lists!
* Can reference data using $ (if the elements are named), or using [], or [[]]

```{r makeList, comment="", prompt=TRUE}
mylist <- list(letters=c("A", "b", "c"), 
        numbers=1:3, matrix(1:25, ncol=5))
```

## List Structure
```{r Lists, comment="", prompt=TRUE}
head(mylist)
```

## List referencing
```{r Listsref1, comment="", prompt=TRUE}
mylist[1] # returns a list
mylist["letters"] # returns a list
```

## List referencing
  
```{r Listsrefvec, comment="", prompt=TRUE}  
mylist[[1]] # returns the vector 'letters'
mylist$letters # returns vector
mylist[["letters"]] # returns the vector 'letters'
```

## List referencing

You can also select multiple lists with the single brackets. 

```{r Listsref2, comment="", prompt=TRUE}
mylist[1:2] # returns a list
```

## List referencing

You can also select down several levels of a list at once

```{r Listsref3, comment="", prompt=TRUE}
mylist$letters[1]
mylist[[2]][1]
mylist[[3]][1:2,1:2]
```

## `%in%`

```{r}
x = c(0, 2, 2, 3, 4)
2 %in% x
5 %in% x
```

```{r}
(x == 0 | x == 2)
x %in% c(0, 2)
```